S-Aligner: Ultrascalable read mapping on Sunway Taihu Light

### 2017-04-07 s-aligner.tex
minor edits throughout

\title{S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light}

\author{\IEEEauthorblockN{Xiaohui Duan, Kai Xu, Yuandong Chan}
\IEEEauthorblockA{%College of Computer Science and Technology \\
Shandong University\\
Jinan, China\\
Email: \{sunrise.duan, xu\_kai, yuandong.c\}@mail.sdu.edu.cn}
\and
\IEEEauthorblockN{Christian Hundt, Bertil Schmidt}
\IEEEauthorblockA{Johannes Gutenberg University\\
Mainz, Germany\\
Email: \{hundt, bertil.schmidt\}@uni-mainz.de}
\and
\IEEEauthorblockN{Pavan Balaji}
\IEEEauthorblockA{Argonne National Laboratory\\
Argonne, Illinois, USA\\
Email: balaji@anl.gov}
\and
\IEEEauthorblockN{Weiguo Liu}
\IEEEauthorblockA{Shandong University\\
Jinan, China\\
Email: weiguo.liu@sdu.edu.cn\\
}
}

@@

>>>>

\title{S-Aligner: Ultrascalable read mapping on Sunway Taihu Light}

\author{\IEEEauthorblockN{Xiaohui Duan, Kai Xu, Yuandong Chan}
\IEEEauthorblockA{%College of Computer Science and Technology \\
Shandong University\\
Jinan, China\\
Email: \{sunrise.duan, xu\_kai, yuandong.c\}@mail.sdu.edu.cn}
\and
\IEEEauthorblockN{Christian Hundt, Bertil Schmidt}
\IEEEauthorblockA{Johannes Gutenberg University\\
Mainz, Germany\\
Email: \{hundt, bertil.schmidt\}@uni-mainz.de}
\and
\IEEEauthorblockN{Pavan Balaji}
\IEEEauthorblockA{Argonne National Laboratory\\
Lemont, Illinois, USA\\
Email: balaji@anl.gov}
\and
\IEEEauthorblockN{Weiguo Liu}
\IEEEauthorblockA{Shandong University\\
Jinan, China\\
Email: weiguo.liu@sdu.edu.cn\\
}
}

@@

\section*{Acknowledgment}
This material is based upon work supported by the U.S. Department of Energy, Office of Science, under contract DE-AC02-06CH11357.

### 2017-04-07 text/abstract.tex
minor edits throughout

  The availability and amount of sequenced genomes have been rapidly
  growing in recent years due to the adoption of next-generation
  sequencing (NGS) technologies that allow for high-throughput
  short-read generation at highly competitive cost. Since this trend
  is expected to continue in the foreseeable future, the design and
  implementation of efficient and scalable NGS bioinformatics
  algorithms are important to research and industrial applications.
  In this paper, we introduce S-Aligner---a highly scalable read
  mapper designed for the \emph{Sunway Taihu Light} supercomputer and
  its fourth-generation ShenWei many-core architecture
  (SW26010). S-Aligner employs a combination of optimization
  techniques to overcome both the memory-bound and the compute-bound
  bottlenecks in the read mapping algorithm. In order to make full use
  of the compute power of Sunway Taihu Light, our design employs three
  levels of parallelism: (1) inter-node parallelism using MPI based on
  a task-grid pattern, (2) intra-node parallelism using
  multi-threading and asynchronous data transfer to fully utilize all
  260 cores of the SW26010 many-core processor, and (3) vectorization
  to exploit the available 256-bit SIMD vector registers.  Moreover,
  we have employed asynchronous access patterns and data sharing
  strategies during file I/O to overcome bandwidth limitations of the
  network file system. Our performance evaluation demonstrates that
  S-Aligner scales almost linearly with approximately 95\% efficiency
  for up to 13,312 nodes (concurrently harnessing more than 3 million
  compute cores). Furthermore, our implementation on a single node
  outperforms the established RazerS3 mapper running on a platform
  with eight Intel Xeon E7-8860v3 CPUs while achieving highly
  competitive alignment accuracy.

>>>>

  The availability and amount of sequenced genomes have been rapidly
  growing in recent years because of the adoption of next-generation
  sequencing (NGS) technologies that emable high-throughput
  short-read generation at highly competitive cost. Since this trend
  is expected to continue in the foreseeable future, the design and
  implementation of efficient and scalable NGS bioinformatics
  algorithms are important to research and industrial applications.
  In this paper, we introduce S-Aligner---a highly scalable read
  mapper designed for the \emph{Sunway Taihu Light} supercomputer and
  its fourth-generation ShenWei many-core architecture
  (SW26010). S-Aligner employs a combination of optimization
  techniques to overcome both the memory-bound and the compute-bound
  bottlenecks in the read mapping algorithm. In order to make full use
  of the compute power of Sunway Taihu Light, our design employs three
  levels of parallelism: (1) internode parallelism using MPI based on
  a task-grid pattern, (2) intranode parallelism using
  multithreading and asynchronous data transfer to fully utilize all
  260 cores of the SW26010 many-core processor, and (3) vectorization
  to exploit the available 256-bit SIMD vector registers.  Moreover,
  we have employed asynchronous access patterns and data-sharing
  strategies during file I/O to overcome bandwidth limitations of the
  network file system. Our performance evaluation demonstrates that
  S-Aligner scales almost linearly with approximately 95\% efficiency
  for up to 13,312 nodes (concurrently harnessing more than 3 million
  compute cores). Furthermore, our implementation on a single node
  outperforms the established RazerS3 mapper running on a platform
  with eight Intel Xeon E7-8860v3 CPUs while achieving highly
  competitive alignment accuracy.

### 2017-04-07 text/background.tex
minor edits throughout

@@

The Sunway Taihu Light has been manufactured by the National Research
Center of Parallel Computer Engineering \& Technology (NRCPC) of the
People's Republic of China and is located at the Wuxi Supercomputing
Center. It provides a theoretical peak performance of 125 PFlop/s and
an effective performance-to-power ratio of over 6 GFlop/s per watt. It
consists of 40,960 nodes with 1.4 PB attached memory. The
interconnection network provides roughly 70 TB/s bisection bandwidth
and 12 GB/s point-to-point communication bandwidth.

@@

Each node is equipped with a single SW26010 processor that is
subdivided into four {\em compute groups} (CGs). A CG is the basic
unit that can be addressed by the job scheduling system. Each CG
consists of a single {\em master processor} (MP), 64 {\em slave
  processors} (SPs), and 8 GB of attached DDR3 shared memory. All
nodes can access a shared network file system for loading and storing
data. The operating system is a customized Linux flavor running on the
MP.  Users may manually launch threads on the SPs in order to
parallelize compute-heavy portions of their code. Both the MP and 64
SPs support ShenWei's RISC basic instruction set, which provides
scalar and SIMD operations. Moreover, reduction primitives can be
called exclusively on the MP, while the SPs exhibit specialized
intrinsics for high-precision integer arithmetic. The 8 GB attached
shared memory can be accessed by both the MP and the 64 SPs via a
memory controller with a shared bandwidth of approximately 136
GB/s. The MP has its own cache (L1 and L2), and each SP can access 64
KB of fast {\em local device memory} (LDM). Data residing in shared
memory can be written to LDM using DMA intrinsics and subsequently
communicated via a broadcast to the LDM of other
SPs. Figure~\ref{fig:SW26010} shows the described hardware layout.

ShenWei's SIMD instruction set features 256-bit wide vector types and
a set of corresponding intrinsics provided by the \emph{ShenWei 5
  Compiler Collection}.  The MPI library (Sunway MPI) is a derivative
of the MPICH implementation, and is compliant with the MPI-3 standard.
Basic threading capabilities for the SPs are provided by the
\emph{athread} library.

@@

We are not the first to parallelize read mapping on a compute
cluster. Among the earliest tools are pBWA \cite{pbwa} and pMap
\cite{pmap}. pBWA is an MPI implementation of Version 0.5.9 of the BWA
aligner while pMap provides an MPI-based framework for the concurrent
execution of popular aligners such as BWA \cite{bwa} and Bowtie
\cite{bowtie}. Other implementations exploit MapReduce frameworks, for
example, the Hadoop-based SEAL \cite{seal} and BigBWA \cite{bigbwa}
tools or the Spark-based SparkBWA suite \cite{sparkBWA}, in order to
guarantee fault tolerance. Unfortunately, all these approaches suffer
from insufficient scalability when executed on hundreds of nodes. The
metaframework parSRA \cite{parSRA} is designed for read mappers
written in UPC++.  It combines dynamic scheduling and a virtual file
system layer for read distribution to overcome the limitations imposed
by pMap and MapReduce-based approaches. merAligner \cite{merAligner}
and CUSHAW3-UPC++ \cite{cushaw3upc} are other PGAS-based short-read
aligners written in UPC; both demonstrate good scalability. Although
pMap and parSRA allow for the execution of single node GPU aligners
such as nvBowtie \cite{nvBio}, none of the cited tools is explicitly
designed for heterogeneous clusters consisting of thousands of
many-core processors. Furthermore, all previous approaches target
traditional hardware architectures.

@@

This problem can be solved by a classical {\em dynamic programming}
(DP) approach that calculates the semi-global alignment between each
$R \in {\cal R}$ and $G$. Unfortunately, each alignment results in a
time complexity proportional to the product of the sequences' lengths,
which renders intractable the alignment of a large number of short
reads to a reference genome a few billion letters long.  In order to
address this problem, most state-of-the art solutions \cite{Reinert}
employ the {\em seed-and-extend} approach, which follows a three-stage
pipeline.

@@

\item[Stage 1: Filtration.] This stage identifies promising candidate
  regions (called {\em seeds}) for each read in $G$. A popular
  strategy in order to discard large regions of $G$ is based on the
  fact that if a read $R \in {\cal R}$ is divided into $e+1$
  non-overlapping $q$-grams (substrings of length $q = \lfloor \lvert
  R\rvert\ /(e+1) \rfloor$), then (according to the pigeonhole
  principle) at least one of them occurs exactly in a match. Such
  occurrences can be identified quickly by looking them up in
  precomputed $q$-gram index data structure (also called the {\em
    reference genome index}), which stores all substrings of length
  $q$ of $G$.

\item[Stage 2. Verification.] This stage determines whether a seed can
  actually be extended to a full match within edit distance $e$. This
  requires the implementation of a {\em verification} algorithm in
  order to analyze the vicinity of each seed. Most mappers typically
  apply fast versions of DP-based algorithms for this step.

\item[Stage 3. Alignment.] This stage generates the base-pair level
  alignment information of a read and its verified intervals in $G$.

@@

Established all-mappers such as RazerS3 \cite{razers3} and mrFAST
\cite{mrfast} follow this pipeline. Our profiling of RazerS3 and
mrFAST using a typical benchmark data set including the human
reference HG19 and 1 million simulated Illumina reads (with a read
length of 100 bps) reveals that Stage 2 occupies 67\% and 93\% of the
overall runtime, respectively. This shows that a fast implementation
of Myers' bit-parallel algorithm for computing the edit distance
between a read and an interval is a crucial component when designing
efficient all-mappers.

Consider two strings $s$ and $s^\prime$ of length $n$ and $m$,
respectively. Their {\em edit distance} is the minimum number of point
mutations (i.e. insertions, deletions, or substitutions) required to
transform $s$ into $s^\prime$. It can be determined by relaxing the
cells of a cost matrix $C$ of size $(n+1) \times (m+1)$ according to
the recurrence relation shown in Equation~\ref{update_scheme}, where
$0 < i \le n$ , $0 < j \le m$ and the {\em characteristic function}
$\chi (x \ne y)$ being 1 if $x \ne y$, and 0 otherwise. Initial
conditions are set to $C[i,0] = i$, $C[0,j] = j$, and $C[0,0] = 0$.

@@

Myers proposed a bit-vector algorithm \cite{myers} that exploits bit
parallelism by encoding the differences (deltas) between adjacent rows
and columns in the cost matrix $C$ as defined in
Equation~\ref{delta_update}, where $\varDelta h_{i,j}$, $\varDelta
v_{i,j}$, and $\varDelta d_{i,j}$ are the discrete derivatives of
$C[i, j]$ in the horizontal, vertical, and diagonal direction,
respectively.

@@

Note that the absolute values in Equation~\ref{delta_update} are
either 0 or 1. Thus, we can encode the relatively small state space
with the help of five vectors using one-hot
encoding. Figure~\ref{BitDP} shows an example for the encoding of the
vertical derivative $\varDelta v$ and its associated one-hot
representations $V^+$ and $V^-$.  With this bit-vector representation
the cell updates can be rewritten in terms of logical operations as
shown in Equation~\ref{one-hot-myers}:

>>>>

This section begins with a description of the Sunlight Taihu Light architecture. Several parallel read mapping approaches are then presented,
before we focus on the seed-and-extend approach

@@

The Sunway Taihu Light has been manufactured by the National Research
Center of Parallel Computer Engineering \& Technology of the
People's Republic of China and is located at the Wuxi Supercomputing
Center. It provides a theoretical peak performance of 125 PFlop/s and
an effective performance-to-power ratio of over 6 GFlop/s per watt. It
consists of 40,960 nodes with 1.4 PB attached memory. The
interconnection network provides roughly 70 TB/s bisection bandwidth
and 12 GB/s point-to-point communication bandwidth.

@@

Each node is equipped with a single SW26010 processor that is
subdivided into four CGs. A CG is the basic
unit that can be addressed by the job scheduling system. Each CG
consists of a single {\em master processor} (MP), 64 {\em slave
  processors} (SPs), and 8 GB of attached DDR3 shared memory. All
%Comment - slave is being replaced by worker  in much of the lit
nodes can access a shared network file system for loading and storing
data. The operating system is a customized Linux flavor running on the
MP.  Users may manually launch threads on the SPs in order to
parallelize compute-heavy portions of their code. Both the MP and 64
SPs support ShenWei's RISC basic instruction set, which provides
scalar and SIMD operations. Moreover, reduction primitives can be
called exclusively on the MP, while the SPs exhibit specialized
intrinsics for high-precision integer arithmetic. The 8 GB attached
shared memory can be accessed by both the MP and the 64 SPs via a
memory controller with a shared bandwidth of approximately 136
GB/s. The MP has its own cache (L1 and L2), and each SP can access 64
KB of fast {\em local device memory} (LDM). Data residing in shared
memory can be written to LDM by using DMA intrinsics and subsequently
communicated via a broadcast to the LDM of other
SPs. Figure~\ref{fig:SW26010} shows the hardware layout.

ShenWei's SIMD instruction set features 256-bit-wide vector types and
a set of corresponding intrinsics provided by the \emph{ShenWei 5
  Compiler Collection}.  The MPI library (Sunway MPI) is a derivative
of the MPICH implementation and is compliant with the MPI-3 standard.
Basic threading capabilities for the SPs are provided by the
\emph{athread} library.

@@

We are not the first to parallelize read mapping on a compute
cluster. Among the earliest tools are pBWA \cite{pbwa} and pMap
\cite{pmap}; pBWA is an MPI implementation of Version 0.5.9 of the BWA
aligner, whereas pMap provides an MPI-based framework for the concurrent
execution of popular aligners such as BWA \cite{bwa} and Bowtie
\cite{bowtie}. Other implementations exploit MapReduce frameworks, for
example, the Hadoop-based SEAL \cite{seal} and BigBWA \cite{bigbwa}
tools and the Spark-based SparkBWA suite \cite{sparkBWA}, in order to
guarantee fault tolerance. Unfortunately, all these approaches suffer
from insufficient scalability when executed on hundreds of nodes. The
metaframework parSRA \cite{parSRA} is designed for read mappers
written in UPC++.  It combines dynamic scheduling and a virtual file
system layer for read distribution to overcome the limitations imposed
by pMap and MapReduce-based approaches.
CUSHAW3-UPC++ \cite{cushaw3upc}
and merAligner \cite{merAligner}
are other PGAS-based short-read
aligners written in UPC; both demonstrate good scalability. Although
pMap and parSRA allow for the execution of single-node GPU aligners
such as nvBowtie \cite{nvBio}, none of the cited tools is explicitly
designed for heterogeneous clusters consisting of thousands of
many-core processors. Furthermore, all previous approaches target
traditional hardware architectures.

@@

This problem can be solved by a classical {\em dynamic programming}
approach that calculates the semi-global alignment between each
$R \in {\cal R}$ and $G$. Unfortunately, each alignment results in a
time complexity proportional to the product of the sequences' lengths,
which renders intractable the alignment of a large number of short
reads to a reference genome a few billion letters long.  In order to
address this problem, most state-of-the-art solutions \cite{Reinert}
employ the {\em seed-and-extend} approach, which follows a three-stage
pipeline.

@@

\item{Stage 1: Filtration.} This stage identifies promising candidate
  regions (called {\em seeds}) for each read in $G$. A popular
  strategy in order to discard large regions of $G$ is based on the
  fact that if a read $R \in {\cal R}$ is divided into $e+1$
  non-overlapping $q$-grams (substrings of length $q = \lfloor \lvert
  R\rvert\ /(e+1) \rfloor$), then (according to the pigeonhole
  principle) at least one of them occurs exactly in a match. Such
  occurrences can be identified quickly by looking them up in
  precomputed $q$-gram index data structure (also called the {\em
    reference genome index}), which stores all substrings of length
  $q$ of $G$.

\item{Stage 2. Verification.} This stage determines whether a seed can
  actually be extended to a full match within edit distance $e$. This
  requires the implementation of a {\em verification} algorithm in
  order to analyze the vicinity of each seed. Most mappers typically
  apply fast versions of DP-based algorithms for this step.

\item{Stage 3. Alignment.} This stage generates the base-pair-level
  alignment information of a read and its verified intervals in $G$.

@@

Established all-mappers such as RazerS3 \cite{razers3} and mrFAST
\cite{mrfast} follow this pipeline. Our profiling of RazerS3 and
mrFAST using a typical benchmark data set including the human
reference HG19 and 1 million simulated Illumina reads (with a read
length of 100 bps) reveals that Stage 2 occupies 67\% and 93\% of the
overall runtime, respectively. These results show that a fast implementation
of Myers' bit-parallel algorithm for computing the edit distance
between a read and an interval is a crucial component when designing
efficient all-mappers.

Consider two strings $s$ and $s^\prime$ of length $n$ and $m$,
respectively. Their {\em edit distance} is the minimum number of point
mutations (i.e., insertions, deletions, or substitutions) required to
transform $s$ into $s^\prime$. It can be determined by relaxing the
cells of a cost matrix $C$ of size $(n+1) \times (m+1)$ according to
the recurrence relation shown in Equation~\ref{update_scheme}, where
$0 < i \le n$ , $0 < j \le m$, and the {\em characteristic function}
$\chi (x \ne y)$ is 1 if $x \ne y$, and 0 otherwise. Initial
conditions are set to $C[i,0] = i$, $C[0,j] = j$, and $C[0,0] = 0$.

@@

Myers proposed a bit-vector algorithm \cite{myers} that exploits bit
parallelism by encoding the differences (deltas) between adjacent rows
and columns in the cost matrix $C$, as defined in
Equation~\ref{delta_update}, where $\varDelta h_{i,j}$, $\varDelta
v_{i,j}$, and $\varDelta d_{i,j}$ are the discrete derivatives of
$C[i, j]$ in the horizontal, vertical, and diagonal direction,
respectively.

@@

Note that the absolute values in Equation~\ref{delta_update} are
either 0 or 1. Thus, we can encode the relatively small state space
with the help of five vectors using one-hot
encoding. Figure~\ref{BitDP} shows an example of the encoding of the
vertical derivative $\varDelta v$ and its associated one-hot
representations $V^+$ and $V^-$.  With this bit-vector representation
the cell updates can be rewritten in terms of logical operations, as
shown in Equation~\ref{one-hot-myers}:

### 2017-04-07 text/conclusion.tex
minor edits throughout

The design of S-Aligner teaches a number of lessons about properly
designing parallelization and communication patterns in order to
achieve both high performance and scalability on the new Sunway Taihu
Light supercomputer. First, both multithreading and SIMD vectorization
need to be employed to fully exploit the computational resources of
the SW26010 processor. Second, within an SP the fast LDM must be used
via DMA intrinsics in order to guarantee efficient intra-CG
communication. Third, a scalable inter-CG communication scheme must be
implemented in a way that efficiently hides the interconnection
network bottleneck. Fourth, asynchronous file-loading and data-sharing
strategies need to be implemented in order to effectively hide the
latency of the network file system. The techniques presented in this
paper could also be adapted to map similar applications exhibiting a
pipeline of hashing and verification, such as large-scale approximate
near duplicate object detection \cite{efficient} or de novo genome
assembly \cite{swap} onto the heterogeneous many-core cluster
architecture of Sunway Taihu Light.

>>>>

From S-Aligner we learn a number of lessons about properly
designing parallelization and communication patterns in order to
achieve both high performance and scalability on the new Sunway Taihu
Light supercomputer. First, both multithreading and SIMD vectorization
must be employed in order to fully exploit the computational resources of
the SW26010 processor. Second, within an SP the fast LDM must be used
via DMA intrinsics in order to guarantee efficient intra-CG
communication. Third, a scalable inter-CG communication scheme must be
implemented in a way that efficiently hides the interconnection
network bottleneck. Fourth, asynchronous file-loading and data-sharing
strategies need to be implemented in order to effectively hide the
latency of the network file system. The techniques presented in this
paper can also be adapted to map similar applications exhibiting a
pipeline of hashing and verification, such as large-scale approximate
near duplicate object detection \cite{efficient} or de novo genome
assembly \cite{swap} onto the heterogeneous many-core cluster
architecture of Sunway Taihu Light.
